Topology, geometry and morphology of the dark matter web

### 2018-06-19 Draft/Chapter4/Paper2017a_Nobmath.tex
conclusion updated
proofed

\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\end{comment}
%-------------------------------------------------------------------------------------------------------------------------------
\chapter{Topology and geometry of the dark matter web}\label{chapter4}

@@

\section{Introduction} 
\label{sec:intro}

@@

Local second order variation in a scalar field $f$ like the multistream field can by found using the Hessian matrix ${\bf H}(f)$. An element of the Hessian matrix is given in \autoref{eq:Hess}, where $i$ and $j$ can be any of $x$, $y$ or $z$ directions.

\begin{equation}
\label{eq:Hess}
 {\bf H}_{ij}(f) = \frac{\partial^2 f}{\partial x_i \partial x_j} 
\end{equation}

@@

\section{Summary}
\label{sec:summary}

>>>>

\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\end{comment}
%-------------------------------------------------------------------------------------------------------------------------------
\chapter{Topology and geometry of dark matter structures}\label{chapter4}

@@

\section{Introduction} 
\label{sec:4intro}

@@

Local second order variation in a scalar field $f$ like the multistream field can by found using the Hessian matrix ${\bf H}(f)$. An element of the Hessian matrix is given in \autoref{eq:aHess}, where $i$ and $j$ can be any of $x$, $y$ or $z$ directions.

\begin{equation}
\label{eq:aHess}
 {\bf H}_{ij}(f) = \frac{\partial^2 f}{\partial x_i \partial x_j} 
\end{equation}

@@

\section{Summary}
\label{sec:4summary}

### 2018-06-19 Draft/thesis-ku.tex
proofed

% 2018 Paper
The cosmic web is one of the most striking features of the distribution of galaxies and dark matter on the largest scales in the Universe. It is composed of dense regions packed full of galaxies, long filamentary bridges, flattened sheets and vast low density voids. The study of the cosmic web has focused primarily on the identification of such features, and on understanding the environmental effects on galaxy formation and halo assembly. As such, a variety of different methods have been devised to classify the cosmic web -- depending on the data at hand, be it numerical simulations, large sky surveys or other. Using various geometrical, topological and heuristic tools in a novel parameter space -- the multistream field -- we explore the distribution of dark matter in cosmological simulations.

@@

I would like to express my deepest gratitute to my adviser Prof. Sergei Shandarin, who has been my source of invaluable advice, encouragement and motivation from the first day of graduate school. Sergei has not only been instrumental in shaping this dissertation, but has also helped me grow as a physicist. His passion for intellectual pursuit and critical thinking will always be my inspiration.

I would also like to thank my thesis committee: Prof. Hume Feldman, Prof. Bruce Twarog, Prof. Barbara Anthony-Twarog, Dr. Suzanne Shontz, Prof. Mikhail Medvedev for supervision and guidance that has been important for the materialisation of this thesis.

My appreciation to the friendly staff and faculty at the Department of Physics and Astronomy, especially Dr. Jennifer Delgado, Kristin Rennells, Kim Hubbel and Joel Saurwein for being kind, supportive and patient with me.

It has been a great pleasure sharing office with Brett, Alex, Yuyu and Keita over the past few years. Stimulating conversations -- physics or otherwise -- was always fun and exciting. Special thanks to Donald and Sinan for the company that kept me sane over the course of PhD.

My time at Lawrence was enjoyable in large part due to many wonderful people I was surrounded by. I've had amazing roommates: Niranjan, Niharika, Darshan and Charu, who

Thank you Mum and Dad, for unconditional love and support throughout graduate school and my life in general.

>>>>

% 2018 Paper

Spatial distribution of dark matter displays variety of intricate three dimensional structures on the largest scales in the Universe, notably the massive haloes, long tubular filaments, flattened walls and the vast under-dense voids. Galaxies embedded in the dark matter structures have illuminated rich geometry of these structures currently known as the cosmic web.

Cosmological N-body simulations are indispensable tools for understanding the evolution of the dark matter web. Recent developments in the numerical analysis of cosmological simulations have hinted towards incorporating the dynamical information of gravitational clustering of collisionless dark matter. This is inferred from a six-dimensional Lagrangian sub-manifold -- comprising of initial and final coordinates of dark matter particles. Velocity multistream field derived from this sub-manifold sheds new light on the nature of gravitational collapse.

Geometrical, topological, morphological and heuristic diagnostic tools used in this novel parameter space reveal features of the dark matter distribution. For instance, a single void structure not only percolates the multistream field in all the directions, but also occupies over 99 per cent of all the single-streaming regions. On the other hand, connected filaments display a rapid topological transition to isolated islands of high multistream values. Hessian analysis delineates structures with different shapes: tubular, sheet-like, or globular -- enabling detection of dark matter haloes without ad hoc parameters related to matter density or distance field.

% 2018 Paper
% The cosmic web is one of the most striking features of the distribution of galaxies and dark matter on the largest scales in the Universe. It is composed of dense regions packed full of galaxies, long filamentary bridges, flattened sheets and vast low density voids. The study of the cosmic web has focused primarily on the identification of such features, and on understanding the environmental effects on galaxy formation and halo assembly. As such, a variety of different methods have been devised to classify the cosmic web -- depending on the data at hand, be it numerical simulations, large sky surveys or other.

@@

I would like to express my deepest gratitude to my adviser Prof. Sergei Shandarin, who has been my source of invaluable advice, encouragement and motivation from the first day of graduate school. Sergei has not only been instrumental in shaping this dissertation, but has also helped me grow as a physicist. His passion for intellectual pursuit and critical thinking will always be my inspiration.

I would also like to thank my thesis committee: Prof. Hume Feldman, Prof. Bruce Twarog, Prof. Barbara Anthony-Twarog, Dr. Suzanne Shontz, Prof. Mikhail Medvedev for supervision and guidance that has been important for the materialization of this thesis.

My appreciation to the friendly faculty and staff at the Department of Physics and Astronomy, especially Kristin Rennells, Kim Hubbel and Joel Sauerwein whose readiness to help students cannot be overestimated.

My time as a graduate student was enjoyable in large part due to many wonderful people I was surrounded by. Anupama has been an inspiring grad student and a wonderful friend, with whom I have shared moments of frustration, laughter and excitement through the entirety of my PhD. I had a great pleasure sharing offices with Brett, Alex, Yuyu and Keita over the past few years. Stimulating conversations with them -- physics or otherwise -- was always fun and exciting. Special thanks to Donald and Sinan for the company that kept me sane over the course of PhD. I've had amazing house mates: Niranjan, Niharika, Darshan and Charu, who have not only put up with my idiosyncrasies, but also occasionally cooked me delicious dinner.

I very much appreciate the opportunities provided by Prof. Salman Habib and the Cosmology group at Argonne, where I have spent better part of this past year. I'm really excited about joining the group as a post-doc in the near future.

Thank you Mum and Dad, for unconditional love and support throughout graduate school and my life in general. They have always encouraged me to pursue my dreams. None of this would have been possible without their care and guidance.
